### 2021-07-10 docs/manuscript/manuscript.tex
Update manuscript.tex

\subsection*{Methods}
In this study we apply a Markov chain Monte Carlo (MCMC)-like algorithm to probe
the design space of three-node GRNs with the ability to generate a band-like
expression pattern (target phenotype) in the middle of an arrangement of 30
cells, which resemble a simple (1-D) morphogenetic field in a developing embryo.
Unlike the majority of modeling studies published so far, here we explore the
space of GRN topologies with nodes having the potential to perceive differently
the same input signal. This allows for a lot more flexibility during the search
space process, and thus enables us to identify a larger set of potentially
interesting and realizable morphogen interpretation mechanisms.

\subsection*{Results}
Out of 2061 GRNs selected using the search space algorithm, we found 714 groups of
network topologies that were able to correctly interpret the morphogen.
Importantly, the main network motif found to generate the target phenotype in
response to the input signal was the type 3 Incoherent Feed Forward Loop (I3-FFL),
which is in agreement with previous theoretical expectations and experimental
observations. Notably, compared to previously reported pattern forming GRN
topologies, we have uncovered a great variety of novel network designs, some of
which might be worth inquiring through synthetic biology methodologies to test
for the ability of network design with minimal regulatory complexity to robustly
interpret a developmental cue.

@@

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step.
% Author Summary not valid for PLOS ONE submissions.
\section*{Author summary}
Systems biology is a fast growing field largely powered by advances in
high-performance computing and sophisticated mathematical modeling of
biological systems. Based on these advances, we are now in a position to not
only mechanistically understand but also to accurately predict the behavior of
complex biological processes, including cell differentiation and spatial pattern
formation during embryogenesis. In this article we use an \textit{in silico}
approach to probe the design space of multi-input, three node Gene Regulatory
Networks (GRNs) capable of generating a striped pattern of gene expression in
the context of a simplified 1-D morphogenetic field.

@@

Cells interact continuously with their environment, which requires very specific
regulatory strategies to avoid potential detrimental responses. Arguably, most
cellular functions arise from the dynamic activity of Gene Regulatory Networks
(GRNs), which play a central role in the interpretation of external and internal
signals. This information
processing function is of particular importance in developmental processes such
as those in which a group of cells differentiate in response to a signalling
molecule. Such molecules were referred to as morphogens by Turing in
1952~\cite{Turing1952} and posterior theoretical studies on patterning led to the
conceptualization of the French Flag Problem by Wolpert, who also stated
in this respect that a gradient of concentration of a morphogen could
trigger cell differentiation in a one-dimensional field of cells~\cite{Wolpert1969,
sharpe_2019}.
Although the Bicoid protein (bcd), the first example of a molecule that acted
as a morphogen, was only found in the 80's in the developing embryos of
\textit{Drosophila melanogaster}~\cite{driever_gradient_1988,
driever_bicoid_1988}, now it is known that there
are many other examples such as the Decapentaplegic (Dpp) protein in
\textit{Drosophila} wing imaginal discs~\cite{affolter_decapentaplegic_2007},
as well as Sonic Hedgehog~\cite{dessaud_dynamic_2010,cohen_theoretical_2014}
and Wnt~\cite{raspopovic_digit_2014}. However, information processing by GRNs
has proven to be a complex process, and a major goal of developmental biology
is to understand mechanistically how positional information conveyed by morphogens
is translated into spatial differentiation.\\

Our understanding of how GRNs process information has increased thanks to the
concept and theory of network motifs, which are defined as patterns of
interconnections occurring in networks at numbers that are significantly higher
than those in randomized networks~\cite{Milo2002}. This novel approach, along
with mathematical modeling and computational systems biology methods have been
shown to be a powerful tool to study GRNs~\cite{kitano_computational_2002}, and
the predictions of studies using these methods have been validated by
experimental studies~\cite{Kalir2005,Kalir2004,Mangan2003a,ODonnell2005}, and
some of them have even used synthetic GRNs to produce artificial cell
differentiation~\cite{Basu2005}, circadian gene
expression~\cite{atkinson_development_2003}, counting
devices~\cite{friedland_synthetic_2009} and systems that respond to light.
Moreover, the design and implementation of GRNs through synthetic biology is
emerging as a promising tool to study biological phenomena as pattern
formation~\cite{santos-moreno_using_2019} and as a novel therapeutic tool with
interesting biomedical applications~\cite{karlsson_therapeutic_2012,
higashikuni_advancing_2017,abil_synthetic_2015,healy_genetic_2019,
kitada_programming_2018}.

In a tissue context, the generation of a stripe of gene expression is a fundamental
patterning function in development, and it has been shown that simple feed-forward
motifs can robustly achieve such a patterning task~\cite{munteanu_2014}. For
example, Cotterell and Sharpe studied  what kind of 3-node network topologies
could effectively translate a morphogen gradient into a striped gene expression
pattern in a one-dimensional field of cells, and found a variety of networks that
implemented at least 7 different mechanisms with varying levels of complexity,
being most of them variations of feed-forward motifs~\cite{Cotterell2010}.

Although GRNs with positional information processing capacities have been
extensively analyzed, most studies have so far placed emphasis on particular
regulatory systems that respond to a single input only. These studies have
typically focused on either regulatory systems observed in
nature~\cite{vonDassow2000,Reinitz1995,Jaeger2004}, synthetic
implementations~\cite{Basu2005,Schaerli2018,Elowitz2000,barbier2020} with
predefined topologies, or have deliberately constrained the study of GRNs to just
a handful of alternative designs~\cite{munteanu_2014,Schaerli2014}. However,
previous work has demonstrated the necessity to expand computational analysis
of pattern forming GRNs to multi-input settings. For instance, it has been shown
that the neural subtypes specification system in the vertebrate neural tube involves
a  two-input network in which Sonic Hedgehog acts as a morphogen and the Olig2
and Nkx2.2 genes can act as the receiver nodes~\cite{dessaud_dynamic_2010,
balaskas2012, exelby_2021}.

In this study, we apply a Markov chain Monte Carlo (MCMC)-like algorithm to
probe the design space of three-node GRNs looking for topologies capable of
translating a morphogen gradient  (input signal) into a striped pattern of gene
expression (phenotype). Importantly, unlike most previous computational
studies, here we allowed any node (e.g. gene) in a GRN to perceive the input
signal in varying ways, and selected the best performing GRNs based on a
fitness criterion used to assess the quality of the phenotype with respect to
a prescribed optimal pattern. Based on this computational strategy, we were
able to uncover a great variety of distinct classes of network topologies that
tend to form a complex interconnected meta-graph that could be easily
traversed throughout evolution via single changes in the wiring of the
different GRNs.

% Lorem ipsum dolor sit~\cite{bib1} amet, consectetur adipiscing elit. Nullam Eq~(\ref{eq:schemeP}) sapien tellus, commodo id velit id, eleifend volutpat quam. Nunc pellentesque pretium diam, quis maximus dolor faucibus id.~\cite{bib2} Nunc convallis sodales ante, ut ullamcorper est egestas vitae.

>>>>

\subsection*{Methods}
In this study, we apply a Markov chain Monte Carlo (MCMC)-like algorithm to probe
the design space of three-node GRNs with the ability to generate a band-like
expression pattern (target phenotype) in the middle of an arrangement of 30
cells, which resemble a simple (1-D) morphogenetic field in a developing embryo.
Unlike most modeling studies published so far, here we explore the
space of GRN topologies with nodes having the potential to perceive differently
the same input signal. This allows for a lot more flexibility during the search
space process, and thus enables us to identify a larger set of potentially
interesting and realizable morphogen interpretation mechanisms.

\subsection*{Results}
Out of 2061 GRNs selected using the search space algorithm, we found 714 classes of
network topologies that could correctly interpret the morphogen.
Notably, the main network motif that generated the target phenotype in
response to the input signal was the type 3 Incoherent Feed-Forward Loop (I3-FFL),
which agrees with previous theoretical expectations and experimental
observations. Particularly, compared to a previously reported pattern forming GRN
topologies, we have uncovered a great variety of novel network designs, some of
which might be worth inquiring through synthetic biology methodologies to test
for the ability of network design with minimal regulatory complexity to 
interpret a developmental cue robustly.

@@

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step.
% Author Summary not valid for PLOS ONE submissions.
\section*{Author summary}
Systems biology is a fast growing field largely powered by advances in
high-performance computing and sophisticated mathematical modeling of
biological systems. Based on these advances, we are now in a position to
mechanistically understand and accurately predict the behavior of
complex biological processes, including cell differentiation and spatial pattern
formation during embryogenesis. In this article, we use an \textit{in silico}
approach to probe the design space of multi-input, three-node Gene Regulatory
Networks (GRNs) capable of generating a striped gene expression pattern in
the context of a simplified 1-D morphogenetic field.

@@

Cells interact continuously with their environment, which requires precise
regulatory strategies to avoid potential detrimental responses. Arguably, most
cellular functions arise from the dynamic activity of Gene Regulatory Networks
(GRNs), which play a central role in interpreting external and internal
signals. This information processing function is critical in 
developmental processes such as those in which a group of cells differentiates in
response to a signaling molecule. Such molecules were referred to as morphogens by Turing in
1952~\cite{Turing1952}, and posterior theoretical studies on patterning led to the
conceptualization of the French Flag Problem by Wolpert, who also stated
in this respect that a gradient of concentration of a morphogen could
trigger cell differentiation in a one-dimensional field of cells~\cite{Wolpert1969,
sharpe_2019}.
Although the Bicoid protein (bcd), the first example of a molecule that acted
as a morphogen, was only found in the 80s in the developing embryos of
\textit{Drosophila melanogaster}~\cite{driever_gradient_1988,
driever_bicoid_1988}, now it is known that there
are many other examples such as the Decapentaplegic (Dpp) protein in
\textit{Drosophila} wing imaginal discs~\cite{affolter_decapentaplegic_2007},
as well as Sonic Hedgehog~\cite{dessaud_dynamic_2010,cohen_theoretical_2014}
and Wnt~\cite{raspopovic_digit_2014}. However, information processing by GRNs
has proven to be a complex process, and a major goal of developmental biology
is to understand mechanistically how positional information conveyed by morphogens
is translated into spatial differentiation.\\

Our understanding of how GRNs process information has increased thanks to the
concept and theory of network motifs, defined as patterns of
interconnections occurring in networks at numbers significantly higher
than those in randomized networks~\cite{Milo2002}. This novel approach, along
with mathematical modeling and computational systems biology methods, is a 
powerful tool to study GRNs~\cite{kitano_computational_2002}, and
experimental studies have validated the predictions of studies using these
methods~\cite{Kalir2005,Kalir2004,Mangan2003a,ODonnell2005}.
Some have even used synthetic GRNs to produce artificial cell
differentiation~\cite{Basu2005}, circadian gene
expression~\cite{atkinson_development_2003}, counting
devices~\cite{friedland_synthetic_2009} and systems that respond to light.
Moreover, the design and implementation of GRNs through synthetic biology is
emerging as a promising tool to study biological phenomena as pattern
formation~\cite{santos-moreno_using_2019} and as a novel therapeutic tool with
interesting biomedical applications~\cite{karlsson_therapeutic_2012,
higashikuni_advancing_2017,abil_synthetic_2015,healy_genetic_2019,
kitada_programming_2018}.

In a tissue context, the generation of a stripe of gene expression is a fundamental
patterning function in development, and it has been shown that simple feed-forward
motifs can robustly achieve such a patterning task~\cite{munteanu_2014}. For
example, Cotterell and Sharpe studied what kind of 3-node network topologies
could effectively translate a morphogen gradient into a striped gene expression
pattern in a one-dimensional field of cells. They found a variety of networks that
implemented at least 7 different mechanisms with varying complexity levels,
most of them variations of feed-forward motifs~\cite{Cotterell2010}.

Although GRNs with positional information processing capacities have been
extensively analyzed, most studies have so far emphasized particular
regulatory systems that respond to a single input only. These studies have
typically focused on either regulatory systems observed in
nature~\cite{vonDassow2000,Reinitz1995,Jaeger2004}, synthetic
implementations~\cite{Basu2005,Schaerli2018,Elowitz2000,barbier2020} with
predefined topologies, or have deliberately constrained the study of GRNs to just
a handful of alternative designs~\cite{munteanu_2014,Schaerli2014}. However,
previous work has demonstrated the necessity to expand computational analysis
of pattern forming GRNs to multi-input settings. For instance, it has been shown
that the neural subtypes specification system in the vertebrate neural tube involves
a  two-input network in which Sonic Hedgehog acts as a morphogen and the Olig2
and Nkx2.2 genes can act as the receiver nodes~\cite{dessaud_dynamic_2010,
balaskas2012, exelby_2021}.

In this study, we apply a Markov chain Monte Carlo (MCMC)-like algorithm to
probe the design space of three-node GRNs looking for topologies capable of
translating a morphogen gradient  (input signal) into a striped pattern of gene
expression (phenotype). Importantly, unlike most previous computational
studies, here we allowed any node (e.g., gene) in a GRN to perceive the input
signal in varying ways and selected the best-performing GRNs based on a
fitness criterion used to assess the quality of the phenotype to
a prescribed optimal pattern. Based on this computational strategy, we uncovered
a great variety of distinct classes of network topologies that
tend to form a complex interconnected meta-graph that could be easily
traversed throughout evolution via single changes in the wiring of the
different GRNs.
